Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-25 Manuscript/SupportingInfo-MLC.Rmd
SI MLC edits

---
title:

output: 
  pdf_document:
    fig_caption: yes
    number_sections: no  #note: change to 'no' for most journals
    keep_tex: true
    pandoc_args:

csl: evolution.csl
bibliography: Lambda-Refs.bib

---
# Supporting Information for: A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

Here we provide additional supporting information referenced in the main document: additional analyses, and simulation results across a wider set of input conditions.

## Simulations on differently shaped phylogenies

@@

In addition to using pure-birth phylogenies, we explored the effect of tree shape on our findings using both balanced and pectinate trees. As before, simulations were conducted on a range of tree sizes ($n=2^5 - 2^{10}$), and across a range of input levels of phylogenetic signal ($\lambda_{in} = 0.0 \to 1.0$; in 21 intervals of 0.05 units). For each $n$ and $\lambda_{in}$ combination, 50 replicates of a continuous trait were simulated using a Brownian motion model of evolution. Using these, we estimated the degree of phylogenetic signal using $\lambda$. \hfill\break

*Results*. As found with pure-birth phylogenies, estimates of $\lambda$ varied most dramatically in simulations with fewer species and at intermediate values of lambda (Fig. S1, S2). Pectinate trees showed an interesting tendency to underestimate $\lambda$ across input values. Most dramatically, some simulations on the largest pectinate phylogeny ($n = 1024$) estimated $\lambda = 0$ for input values as high as 0.5 (Fig. S2).

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS1.png")

```

\textbf{Figure S1}. Accuracy of Pagel's lambda estimations across known lambda inputs on various tree sizes. Results obtained using balanced trees. \hfill\break

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS2.png")

```

\textbf{Figure S2}. Accuracy of Pagel's lambda estimations across known lambda inputs on various tree sizes. Results obtained using pectinate trees.

## Simulations of phylogenetic regression and ANOVA

We analyzed the precision of parameter estimation when parameters were obtained in phylogenetic regression and ANOVA. This involved simulations of dependent variables with $\lambda$ values ranging from 0 to 1 (21 intervals of 0.05 units) across pure-birth trees ($n=2^5 - 2^{10}$). For each $\lambda$ input value, independent variables were then generated with a known relationship to the simulated dependent variable ($\beta = 0, 0.25, 0.5, 0.75, 1.0$), from which $\lambda$ was then estimated.

*Results* As with the results from the regression analyses in the main text, we found poor precision of $\lambda$ estimation in smaller trees and at intermediate levels of phylogenetic signal with phylogenetic ANOVA (Fig. S3). Slope was reliably estimated across tree sizes and known $\lambda$ values (Fig. S4).

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS3.png")

```

\textbf{Figure S3}. Precision of Pagel's $\lambda$ when incorporated in phylogenetic ANOVA. Results are from input $\beta$ values of $0.5$. \hfill\break

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS4.png")

```

\textbf{Figure S4}. Precision of slope estimates across known input $\beta$ values of phylogenetic regression. Results shown include all input $\lambda$ values.

## Effect Sizes of Pagel's Lambda ($Z_\lambda$) and Kappa ($Z_K$)

We proposed the use of effect sizes (Z-scores) as a measure of the strength of phylogenetic signal. These were obtained from summary parameters of phylogenetic signal: $\lambda$ and $Kappa$. The precision of effect sizes $Z_\lambda$ and $Z_K$ are summarized in the main text, primarily for simulations where $n=32$. Here we present results from additional simulations.

*Results* Effect sizes for Pagel's $\lambda$ ($Z_\lambda$) scale nonlinearly with input levels of phylogenetic signal, and this pattern is consistent across all tree sizes examined (Figure S5). Further, the precision of $Z_\lambda$ varied widely across input levels of phylogenetic signal, with decidely greater variation (less precision) at stronger input levels of phylogenetic signal (Figure S5). By contrast, effect sizes for *Kappa* ($Z_K$) displayed a linear increase with increasing input levels of phylogenetic signal, which remained consistent across all tree sizes examined (Figure S6). Additionally, the precision of $Z_K$ remained constant across input levels of phylogenetic signal (Figure S6). Together these results imply that $Z_K$ is a more reliable estimate of the strength of phylogenetic signal.

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS5.png")

```

\textbf{Figure S5}. Variation in effect size (Z score) of Pagel's $\lambda$ across input $\lambda$ values on various tree sizes.  \hfill\break

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS6.png")

```

\textbf{Figure S6}. Variation in effect size (Z score) of $K$ across input $\lambda$ values on various tree sizes. \hfill\break

```{r, echo = FALSE, out.width="95%"}
include_graphics("FigS7.png")

```

\textbf{Figure S7}. Estimates of phylogenetic signal using $Kappa$ across known lambda inputs on various tree sizes.  Plot shows that while $Kappa$ increases with increasing phylogenetic signal, the precision of this estimate also varies. Thus, conversion to a standardized effect size ($Z_K$) is preferred.

>>>>

---
title: "Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal"

author: "Dean C. Adams, Erica K. Baken, and Michael L. Collyer"

output: 
  pdf_document:
    fig_caption: yes
    keep_tex: true
    pandoc_args:

csl: pnas.csl
bibliography: Lambda-Refs.bib
---

@@

Here we provide additional supporting information referenced in the main document, which include additional analyses, and simulation results across a wider set of input conditions. As before, simulations were conducted on six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 50 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 50 trees we measured the mean values of $\hat{\lambda}$ and $K$, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness.

## Additional Results on Pure-Birth Phylogenies

Results from the simulations on pure-birth trees revealed that both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size, whereas $Z_K$ was more consistent (Fig. S4). Additionally, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, and its standard deviation across input signal was more even across tree sizes, implying more consistent precision. This latter observation is seen in the upper four panels of Fig. S1, where the dispersion of simulation results (gray dots) is far more even at each input level of phylogenetic signal for $Z_K$ than it is for $Z_{\lambda}$.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S1.png")  
```

\textbf{Figure S1}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion for pure-birth trees.

\newpage

## Simulations on Pectinate Phylogenies

Results from simulations on pectinate phylogenies largely mirrored those found on pure-birth trees. For $\hat{\lambda}$, the mean value increased with increasing input signal, but was negatively biased, and was less than the input value across most of its range (Fig. S2: black line). Additionally, the precision of $\hat{\lambda}$ varied with differing input levels, with the greatest variation found at intermediate values of $\lambda$ (Fig. S2 red line). Finally, the distribution of $\hat{\lambda}$ was not normal, and became more skewed at more extreme values of $\lambda$ (Fig. S2 blue line).  \hfill\break

For *K*, mean values increased with increasing phylogenetic signal, though as was found with pure-birth trees, the increase was nonlinear (Fig. S3 black line). Likewise, variation increased with increasing phylogenetic signal (Fig. S3 red line), though the distribution of *K* was more normally distributed throughout its range, and across different tree sizes, as compared with $\hat{\lambda}$, though there was some slight skew for high input levels of phylogenetic signal on large phylogenies (Fig. S3 blue line). \hfill\break

With respect to effect sizes, both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size, whereas $Z_K$ was more consistent (Fig. S4). Also, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, and its standard deviation across input signal was more even across tree sizes, implying more consistent precision.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S2.png")  
```

\textbf{Figure S2}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion on pectinate trees. Gray line signifies the 1:1 line where the input value matches the estimate.At each input level, the dark
black line represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S3.png")  
```

\textbf{Figure S3}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion on pectinate trees.  At each input level, the dark black line represents the empirically derived expected value (mean) of
\textit{K}, the red line is the standard deviation of
\textit{K}, and the blue line is Shapiro Wilks statistic of
\textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S4.png")  
```

\textbf{Figure S4}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion for pectinate trees.

\newpage

## Simulations on Balanced Phylogenies

Results from simulations on balanced phylogenies also mirrored those found on pure-birth trees. For $\hat{\lambda}$, the mean value increased with increasing input signal, but was negatively biased, and was less than the input value across most of its range (Fig. S5: black line). Additionally, the precision of $\hat{\lambda}$ varied with differing input levels, with the greatest variation found at intermediate values of $\lambda$ (Fig. S5 red line). Finally, the distribution of $\hat{\lambda}$ was not normal, and became more skewed at more extreme values of $\lambda$ (Fig. S5 blue line).  \hfill\break

For *K*, mean values increased with increasing phylogenetic signal, though as was found with pure-birth trees, the increase was nonlinear (Fig. S6 black line). Likewise, variation increased with increasing phylogenetic signal (Fig. S6 red line), though the distribution of *K* was more normally distributed throughout its range, and across different tree sizes, as compared with $\hat{\lambda}$, though there was some slight skew for high input levels of phylogenetic signal on large phylogenies (Fig. S6 blue line). \hfill\break

With respect to effect sizes, both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size (fig. S7). Also, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, though there was a slight effect of tree size with balanced phylogenies. But as before, its standard deviation across input signal was more even across tree sizes, implying more consistent precision.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S5.png")  
```

\textbf{Figure S5}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion on balanced trees. Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S6.png")  
```

\textbf{Figure S6}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion on balanced trees.  At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S7.png")  
```

\textbf{Figure S7}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion for balanced trees.

\newpage

## Simulations on Phylogenies Containing Polytomies

We also investigaged the effect of unresolved phylogenies on estimates of phylogenetic signal by adjusting pure-birth trees to have 20% collapsed nodes, following the procedures of [@MolinaVenegas2017]. Results from simulations on phylogenies containing polytomies mirrored those found on fully resolved trees. For $\hat{\lambda}$, the mean value increased with increasing input signal, but was negatively biased, and was less than the input value across most of its range (Fig. S8: black line). Additionally, the precision of $\hat{\lambda}$ varied with differing input levels, with the greatest variation found at intermediate values of $\lambda$ (Fig. S8 red line). Finally, the distribution of $\hat{\lambda}$ was not normal, and became more skewed at more extreme values of $\lambda$ (Fig. S8 blue line).  \hfill\break

For *K*, mean values increased with increasing phylogenetic signal, though as was found with pure-birth trees, the increase was nonlinear (Fig. S9 black line). Likewise, variation increased with increasing phylogenetic signal (Fig. S9 red line), though the distribution of *K* was more normally distributed throughout its range, and across different tree sizes, as compared with $\hat{\lambda}$ (Fig. S9 blue line). \hfill\break

With respect to effect sizes, both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size, whereas $Z_K$ was more consistent (Fig. S10). Also, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, and its standard deviation across input signal was more even across tree sizes, implying more consistent precision. Thus, polytomies do not exert an appreciable effect on $Z_K$.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S8.png")  
```

\textbf{Figure S8}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion on pure-birth trees containing polytomies. Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S9.png")  
```

\textbf{Figure S9}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion on pure-birth trees containing polytomies.  At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S10.png")  
```

\textbf{Figure S10}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion for pure-birth trees containing polytomies.

\newpage

## Simulations of Phylogenetic Regression and ANOVA

We also investigated the effect of incorporating $\lambda$ in PGLS analyses (phylogenetic regression and ANOVA). Here, patterns for $\hat\lambda$, *K*, $Z_\lambda$ and $Z_K$ were virtually identical to those found previously. In essence, $\hat\lambda$ was not suitable as a test statistic representing phylogenetic signal due to all properties previously shown (Figs. S11 & S12), *K* was more appropriate, but  associated non-linearly with signal strength (Figs. S13 & 14), while $Z_K$ was found to be a superior effect size as compared with $Z_\lambda$ (Figs S15 & S16). \hfill\break

With respect to model performance, input parameters ($\beta$) were well estimated in phylogenetic regression  and ANOVA when $\lambda$ was incorporated (Figs S17 & S18), implying that $\lambda$ may be used as a tuning parameter for these models. Additionally, type I error and statistical power were unaffected by the inclusion of $\lambda$ (Figs S19 & S20), though power was reduced at smaller sample sizes in phylogenetic ANOVA, mirroring prior findings of [@AdamsCollyer2018b]. These results generally confirmed earlier findings of [@Revell2010] for phylogenetic regression, and extend them to the case of phylogenetic ANOVA.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S11.png")  
```

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S12.png")  
```

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S13.png")  
```

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$).  At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S14.png")  
```

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S15.png")  
```

\textbf{Figure S15}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S16.png")  
```

\textbf{Figure S16}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S17.png")  
```

\textbf{Figure S17}. Parameter estimates from phylogenetic regression (incorporating $\lambda$). The red line represents the estimated slope, which almost completely overlaps the gray 1:1 line.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S18.png")  
```

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA (incorporating $\lambda$). The red line represents the estimated slope, which almost completely overlaps the gray 1:1 line.

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S19.png")  
```

\textbf{Figure S19}. Power curves for phylogenetic regression (incorporating $\lambda$). Type I error is found on the far-left of the plot (effect = 0).

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S20.png")  
```

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating $\lambda$). Type I error is found on the far-left of the plot (effect = 0).

\newpage

## References
